D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-01-29 06related_conclusion.tex

The most obvious difference is that D-Hammer targets labelled Dirac
notation, whereas DiracDec targets plain Dirac notation. As already
mentioned, the use of labelled Dirac notation is essential in many
applications; in particular, labelled Dirac notation can simplify
notation and proofs, and is in general better suited for writing and
reasoning about complex, many-body, quantum states. However, there are
several other important differences. First, our language leverages the
power of higher-order dependent type theory. In particular, we
leverage higher-order functions to provide a compact and expressive
representation of big operators. Second, our language supports AC
symbols with indefinite arities, which leads to compact representation
of terms, and eases AC reasoning. Third, the relation with Mathematica
is fundamentally different. DiracDec is implemented in Mathematica; as
a consequence, its behavior, and in particular, its typing rules are
constrained by the type system of Mathematica. In contrast, D-Hammer
is designed as a separate tool; as a consequence, it benefits from an
improved representation of terms (e.g.\, AX operators with indefinite
arity), a more expressive type system (e.g.\, dependent types), and a
more efficient rewriting engine. D-Hammer still relies on Mathematica
to reason about functions that are not natively supported by rewriting
rules, but these interactions are constrained and do not have negative
effects on the overall efficiency of the system. This is reflected in
our experimental comparison of D-Hammer and DiracDec, which shows how
the former outperforms the latter.

>>>>

The most obvious difference is that D-Hammer targets labelled Dirac
notation, whereas DiracDec targets plain Dirac notation. As already
mentioned, the use of labelled Dirac notation is essential in many
applications; in particular, labelled Dirac notation can simplify
notation and proofs, and is in general better suited for writing and
reasoning about complex, many-body, quantum states. However, there are
several other important differences. First, our language leverages the
power of higher-order dependent type theory. In particular, we
leverage higher-order functions to provide a compact and expressive
representation of big operators. Second, our language supports AC
symbols with indefinite arities, which leads to compact representation
of terms, and eases AC reasoning. Third, the relation with Mathematica
is fundamentally different. DiracDec is implemented in Mathematica; as
a consequence, its behavior, and in particular, its typing rules are
constrained by the type system of Mathematica. In contrast, D-Hammer
is designed as a separate tool; as a consequence, it benefits from an
improved representation of terms (e.g.\, AC operators with indefinite
arity), a more expressive type system (e.g.\, dependent types), and a
more efficient rewriting engine. D-Hammer still relies on Mathematica
to reason about functions that are not natively supported by rewriting
rules, but these interactions are constrained and do not have negative
effects on the overall efficiency of the system. This is reflected in
our experimental comparison of D-Hammer and DiracDec, which shows how
the former outperforms the latter.
